---
title: "Getting started with colockit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with colockit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

colockit provides a two-stage pipeline for genomic colocalisation analysis:

1. **`prepare_regions()`** — harmonises two GWAS datasets within a locus,
   aligns effect alleles, and computes an LD matrix from a PLINK reference panel.
2. **`colocalise()`** — runs the chosen colocalisation method (ABF, SuSiE, or
   PWCoCo) on the prepared regions.

Two backends are available:

- **`"standalone"`** — no external database required; you supply GWAS summary
  statistics directly as `data.table`s.
- **`"genoscores"`** — integrates with the Genoscores database to retrieve
  published GWAS or generate private GWAS via SNPTEST.

---

## Input format (standalone backend)

Both `gwas1` and `gwas2` must be `data.table`s with the following columns:

| Column | Type | Description |
|---|---|---|
| `snp` | character | rsid, e.g. `"rs997567"` |
| `dbsnpid` | integer | numeric dbSNP identifier |
| `chrom` | integer | chromosome number |
| `pos` | integer | genomic position (hg38) |
| `allele` | character | effect allele |
| `oth.allele` | character | other allele |
| `beta` | numeric | effect size |
| `pvalue` | numeric | association p-value |
| `n_samples` | integer | sample size |
| `freq` | numeric | effect allele frequency |
| `trait.name` | character | trait label |

---

## Standalone flow

```{r standalone, eval=FALSE}
library(colockit)
library(data.table)

## Define the locus of interest (hg38 coordinates)
locus <- data.table(
  chrom    = 22L,
  startpos = 29001000L,
  endpos   = 29011000L
)

## gwas1 and gwas2 are data.tables in the format described above.
## Here we assume they have been loaded from summary statistic files.
gwas1 <- fread("trait1_summary_stats.csv")
gwas2 <- fread("trait2_summary_stats.csv")

## Stage 1 — harmonise, compute LD
regions <- prepare_regions(
  locus        = locus,
  gwas1        = gwas1,
  gwas2        = gwas2,
  output_dir   = "results/locus_22_29Mb",
  refplinkfile = "path/to/kg.2020.hg38.eur",
  backend      = "standalone"
)

## Stage 2 — run colocalisation
## ABF (fast, single causal variant assumption)
result_abf <- colocalise(
  regions       = regions,
  locus         = locus,
  analysis_type = "abf",
  output_dir    = "results/locus_22_29Mb"
)
result_abf$summary

## SuSiE (handles multiple causal variants)
result_susie <- colocalise(
  regions       = regions,
  locus         = locus,
  analysis_type = "susie",
  output_dir    = "results/locus_22_29Mb"
)
result_susie$summary
```

`prepare_regions()` writes the following files to `output_dir`:

- `gwas1.csv`, `gwas2.csv` — harmonised summary statistics
- `ld.matrix.Rdata.gz` — LD correlation matrix
- `locus.plot.png` — preview Manhattan-style locus plot

`colocalise()` additionally writes:

- `formatted.region.Rdata.gz` — formatted input for the coloc method
- `coloc.abf.png` / `coloc.susie.png` — colocalisation plot with LD heatmap

---

## Genoscores flow

The genoscores backend works identically from the user's perspective,
but retrieves GWAS data from the Genoscores database.

```{r genoscores, eval=FALSE}
library(DBI)

## Establish a database connection (details depend on your installation)
con <- dbConnect(...)

## Published GWAS: supply integer gwasids
regions <- prepare_regions(
  locus        = locus,
  gwas1        = 42L,   # gwasid for trait 1
  gwas2        = 57L,   # gwasid for trait 2
  output_dir   = "results/locus_22_29Mb",
  refplinkfile = "path/to/kg.2020.hg38.eur",
  backend      = "genoscores",
  gwas_type    = "published",
  con          = con
)

## Mixed: published GWAS + private GWAS
private_gwas <- create_private_gwas(
  outcome    = "bmi",
  samplefile = "samples.sample",
  plinkfile  = "genotypes/all_chr",
  locus      = locus,
  output_dir = "results/private_gwas"
)

regions_mixed <- prepare_regions(
  locus        = locus,
  gwas1        = 42L,
  gwas2        = private_gwas,
  output_dir   = "results/locus_22_29Mb_mixed",
  refplinkfile = "path/to/kg.2020.hg38.eur",
  backend      = "genoscores",
  gwas_type    = "mixed",
  con          = con
)
```

The genoscores backend additionally:

- Resolves missing `oth.allele` values by querying the Genoscores dbSNP table
- Retrieves effect allele frequencies from the `r_freqs` table
- Cross-checks SNPTEST-derived frequencies against dbSNP (for private GWAS)

---

## References

- Robinson JW, Hemani G, Babaei MS, Huang Y, Baird DA, Tsai EA, Chen C-Y,
  Gaunt TR, Zheng J (2022). An efficient and robust tool for colocalisation:
  Pair-wise Conditional and Colocalisation (PWCoCo). *bioRxiv*.
  <https://doi.org/10.1101/2022.08.08.503158>

---

## Interpreting results

The `$summary` vector from `coloc.abf` contains posterior probabilities for
five hypotheses:

| Hypothesis | Interpretation |
|---|---|
| H0 | Neither trait has a genetic association in this region |
| H1 | Only trait 1 is associated |
| H2 | Only trait 2 is associated |
| H3 | Both traits are associated, but with **different** causal variants |
| H4 | Both traits are associated, sharing a **single** causal variant |

A value of PP.H4 > 0.8 is commonly used as evidence for colocalisation.

See `vignette("simulated-colocalisation")` for a worked example where
colocalisation is known from the simulation design.
