---
title: "Simulated colocalisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulated colocalisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why simulation?

In a real analysis you never know the ground truth. Simulation lets us
verify that the pipeline gives the correct answer when we *do* know it,
providing confidence before applying it to real data.

The `coloc` package ships a built-in dataset (`coloc_test_data`) with four
simulated GWAS traits and a known causal structure:

| Dataset | Causal SNP(s) |
|---|---|
| D1 | SNP 105 |
| D2 | SNP 105 |
| D3 | SNPs 105 and 89 |
| D4 | SNP 105 |

Because D1 and D2 share the same causal variant (SNP 105), we expect
colocalisation analysis to return a high PP.H4. The LD matrix is
pre-computed from a simulated reference panel.

---

## Preparing the `regions` object

`colocalise()` takes a named list `list(src, trgt, ldmat)` as returned by
`prepare_regions()`. When the LD matrix is already available (as here), we can
construct this list directly without running `prepare_regions()`.

```{r prepare}
library(colockit)
library(data.table)

## Load the coloc test dataset
data("coloc_test_data", package = "coloc")
D1 <- coloc_test_data$D1
D2 <- coloc_test_data$D2

cat("D1 causal SNP:", coloc_test_data$causals$D1, "\n")
cat("D2 causal SNP:", coloc_test_data$causals$D2, "\n")
```

```{r build-regions}
## Helper: convert a coloc-format list to a colockit-compatible data.table
coloc_to_dt <- function(d, trait_name) {
  data.table(
    snp        = d$snp,
    dbsnpid    = as.integer(sub("^s", "", d$snp)),
    chrom      = 1L,
    pos        = as.integer(d$position),
    allele     = "A",
    oth.allele = "G",
    beta       = d$beta,
    varbeta    = d$varbeta,
    se         = sqrt(d$varbeta),
    pvalue     = 2 * pnorm(-abs(d$beta / sqrt(d$varbeta))),
    freq       = d$MAF,
    n_samples  = as.integer(d$N),
    trait.name = trait_name
  )
}

src  <- coloc_to_dt(D1, "Trait 1 (causal: SNP 105)")
trgt <- coloc_to_dt(D2, "Trait 2 (causal: SNP 105)")

## Use the LD matrix included in the test data
ldmat <- D1$LD
rownames(ldmat) <- D1$snp
colnames(ldmat) <- D1$snp

regions <- list(src = src, trgt = trgt, ldmat = ldmat)

cat(nrow(src), "SNPs, LD matrix", nrow(ldmat), "x", ncol(ldmat), "\n")
```

---

## Running ABF colocalisation

```{r colocalise-abf, message=FALSE}
out_dir <- file.path(tempdir(), "simulated_coloc")

result <- colocalise(
  regions       = regions,
  locus         = data.table(chrom = 1L, startpos = 0L, endpos = 501L),
  analysis_type = "abf",
  output_dir    = out_dir
)
```

### Posterior probabilities

```{r posteriors}
pp <- result$summary[c("PP.H0.abf","PP.H1.abf","PP.H2.abf",
                        "PP.H3.abf","PP.H4.abf")]
round(pp, 4)
```

PP.H4 is very close to 1 — strong evidence that both traits share a single
causal variant, consistent with the simulation design.

---

## Contrasting with distinct causal variants

D1 (causal: SNP 105) and D3 (causals: SNPs 105 **and** 89) share one causal
variant but D3 has an additional signal. We expect PP.H4 to remain high under
ABF (which assumes a single causal variant per trait), but SuSiE would
distinguish the extra signal in D3.

```{r contrast}
D3    <- coloc_test_data$D3
trgt3 <- coloc_to_dt(D3, "Trait 3 (causals: SNPs 105 & 89)")
ldmat3 <- D3$LD
rownames(ldmat3) <- D3$snp
colnames(ldmat3) <- D3$snp

regions13 <- list(src = src, trgt = trgt3, ldmat = ldmat3)

out_dir13 <- file.path(tempdir(), "simulated_coloc_13")
result13 <- colocalise(
  regions       = regions13,
  locus         = data.table(chrom = 1L, startpos = 0L, endpos = 501L),
  analysis_type = "abf",
  output_dir    = out_dir13
)

pp13 <- result13$summary[c("PP.H0.abf","PP.H1.abf","PP.H2.abf",
                             "PP.H3.abf","PP.H4.abf")]
round(pp13, 4)
```

ABF still returns high PP.H4 here because the shared causal (SNP 105)
dominates the signal and ABF cannot model multiple causal variants per trait.
For loci where multiple independent signals are suspected, use
`analysis_type = "susie"`.

---

## Running SuSiE colocalisation

SuSiE fits a sum-of-single-effects model that can accommodate multiple causal
variants per trait.

```{r colocalise-susie, message=FALSE, warning=FALSE}
out_dir_susie <- file.path(tempdir(), "simulated_coloc_susie")

result_susie <- colocalise(
  regions       = regions,
  locus         = data.table(chrom = 1L, startpos = 0L, endpos = 501L),
  analysis_type = "susie",
  output_dir    = out_dir_susie
)

if (!is.null(result_susie)) {
  cat("SuSiE summary (best credible set):\n")
  print(result_susie$summary)
}
```

---

## Key take-aways

- When the causal variant is truly shared, colockit correctly identifies
  PP.H4 ≈ 1 using ABF.
- ABF assumes a single causal variant per region; use SuSiE for loci with
  multiple independent signals.
- The `colocalise()` function saves a colocalisation plot (`coloc.abf.png` /
  `coloc.susie.png`) alongside the posterior probabilities, colouring each
  SNP by PP.H4 and labelling the top variant.

For a real-data workflow, see `vignette("getting-started")`.
